---
title: "Working with GridMET Data for California Counties"
subtitle: "A Tutorial Using the Amadeus Package"
author: "Amadeus Package Tutorial"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    code-tools: true
    embed-resources: true
    theme: cosmo
    highlight-style: github
execute:
  warning: false
  message: false
---

## Introduction

This tutorial demonstrates how to use the **amadeus** R package to download, process, and calculate gridMET meteorological data for California counties. gridMET is a high-resolution (~4-km) gridded surface meteorological dataset covering the contiguous United States from 1979-present.

### What You'll Learn

- How to download gridMET data from the Climatology Lab
- How to process raw netCDF files into usable spatial formats
- How to extract meteorological values for California counties
- How to visualize and analyze the results

### Prerequisites

```{r setup}
# Load required packages
library(amadeus)
library(terra)
library(sf)
library(dplyr)
library(ggplot2)
library(tidyr)

# Set random seed for reproducibility
set.seed(123)
```

## Step 1: Obtain California County Boundaries

First, we'll get the spatial boundaries for California counties. We'll use the `tigris` package to download county boundaries from the US Census Bureau.

```{r get-counties}
# Install tigris if not already installed
if (!require(tigris)) install.packages("tigris")
library(tigris)

# Download California counties
# Set options to cache the data
options(tigris_use_cache = TRUE)
ca_counties <- counties(state = "CA", cb = TRUE, progress_bar = FALSE)

# Convert to sf object and select key columns
ca_counties_sf <- st_as_sf(ca_counties) %>%
  select(GEOID, NAME, ALAND) %>%
  rename(county_name = NAME, 
         land_area = ALAND)

# Display first few counties
head(ca_counties_sf)
```

```{r plot-counties}
# Visualize California counties
ggplot(ca_counties_sf) +
  geom_sf(fill = "lightblue", color = "white", size = 0.3) +
  theme_minimal() +
  labs(title = "California Counties",
       subtitle = paste(nrow(ca_counties_sf), "counties"))
```

## Step 2: Download gridMET Data

Now we'll download gridMET data. For this example, we'll download maximum temperature (tmmx) and precipitation (pr) for a specific time period.

```{r setup-download, eval = TRUE}
# Create a temporary directory for data storage
# In practice, you'd want to use a permanent location
print(getwd())
data_dir <- file.path("tutorials", "gridmet_data")
dir.create(data_dir, showWarnings = FALSE, recursive = TRUE)

print(paste("Data will be saved to:", data_dir))
```

### Download Maximum Temperature

```{r download-tmmx, eval=TRUE}
# Download maximum temperature data
# Note: This downloads data for 2023
download_data(
  dataset_name = "gridmet",
  variable = "tmmx",
  year = 2023,
  directory_to_save = data_dir,
  acknowledgement = TRUE,
  download = TRUE,
  remove_command = TRUE,
  hash = FALSE
)
```

::: {.callout-note}
## Note on Data Download

The code above is set to `eval=TRUE` because downloading gridMET data can take several minutes and requires significant storage space (~1-2 GB per variable per year). When running this tutorial yourself, set `eval=TRUE` to actually download the data.
:::

### Download Precipitation

```{r download-pr, eval=TRUE}
# Download precipitation data
download_data(
  dataset_name = "gridmet",
  variable = "pr",
  year = 2023,
  directory_to_save = data_dir,
  acknowledgement = TRUE,
  download = TRUE,
  remove_command = TRUE,
  hash = FALSE
)
```

## Step 3: Process gridMET Data

After downloading, we need to process the raw netCDF files into `SpatRaster` objects that R can work with efficiently.

```{r process-data, eval=TRUE}
# Process maximum temperature data
# Let's focus on summer months (June-August 2023)
tmmx_processed <- process_covariates(
  covariate = "gridmet",
  variable = "tmmx",
  date = c("2023-06-01", "2023-08-31"),
  path = file.path(data_dir, "tmmx")
)

# Process precipitation data
pr_processed <- process_covariates(
  covariate = "gridmet",
  variable = "pr",
  date = c("2023-06-01", "2023-08-31"),
  path = file.path(data_dir, "pr")
)

# View the processed data
print("Maximum Temperature SpatRaster:")
print(tmmx_processed)

print("Precipitation SpatRaster:")
print(pr_processed)
```

::: {.callout-tip}
## Understanding SpatRaster

The `SpatRaster` object contains multiple layers - one for each day in our date range. The `terra` package provides powerful tools for working with these multi-layer raster datasets.
:::

## Step 4: Extract Values at County Locations

Now we'll extract the gridMET values for our counties using their polygon boundaries.

```{r calculate-values, eval=TRUE}
# Extract maximum temperature values using county polygons
# This calculates zonal statistics across the entire county area
tmmx_counties <- calculate_gridmet(
  from = tmmx_processed,
  locs = ca_counties_sf,
  locs_id = "GEOID",
  radius = 0,
  fun = "mean",
  geom = FALSE
)

# Extract precipitation values using county polygons
pr_counties <- calculate_gridmet(
  from = pr_processed,
  locs = ca_counties_sf,
  locs_id = "GEOID",
  radius = 0,
  fun = "mean",
  geom = FALSE
)

# View results
head(tmmx_counties)
head(pr_counties)
```

::: {.callout-note}
## Polygon-based Extraction

By passing the polygon geometries directly (instead of centroids), `calculate_gridmet()` performs **zonal statistics** - calculating the mean across all grid cells within each county boundary. This provides a much more accurate representation than point extraction, especially for large counties.
:::

## Step 5: Data Analysis and Visualization

### Reshape and Summarize Data

```{r analyze-data, eval=TRUE}
# Combine temperature and precipitation data
# Note: calculate_gridmet returns data in long format with a 'time' column
# and variable columns (e.g., 'tmmx_0' for temperature at 0m radius)

# Check the structure
cat("Temperature data structure:\n")
str(tmmx_counties)
cat("\nPrecipitation data structure:\n")
str(pr_counties)

# Convert time column to Date and rename variable columns
tmmx_long <- tmmx_counties %>%
  mutate(date = as.Date(time)) %>%
  rename(tmax_k = tmmx_0) %>%
  mutate(tmax_c = tmax_k - 273.15) %>%  # Convert Kelvin to Celsius
  select(-time, -tmax_k)

pr_long <- pr_counties %>%
  mutate(date = as.Date(time)) %>%
  rename(precip_mm = pr_0) %>%
  select(-time)

# Merge datasets
weather_data <- tmmx_long %>%
  left_join(pr_long, by = c("GEOID", "date")) %>%
  left_join(
    ca_counties_sf %>% st_drop_geometry() %>% select(GEOID, county_name),
    by = "GEOID"
  )

# Calculate summer statistics by county
summer_stats <- weather_data %>%
  group_by(GEOID, county_name) %>%
  summarize(
    mean_tmax_c = mean(tmax_c, na.rm = TRUE),
    max_tmax_c = max(tmax_c, na.rm = TRUE),
    total_precip_mm = sum(precip_mm, na.rm = TRUE),
    days_over_35c = sum(tmax_c > 35, na.rm = TRUE),
    .groups = "drop"
  )

# View hottest counties
summer_stats %>%
  arrange(desc(mean_tmax_c)) %>%
  head(10)
```

### Visualize Temperature Patterns

```{r plot-temperature, eval=TRUE}
# Join summer statistics with county geometry for mapping
ca_summer_map <- ca_counties_sf %>%
  left_join(summer_stats, by = "GEOID")

# Map mean maximum temperature
ggplot(ca_summer_map) +
  geom_sf(aes(fill = mean_tmax_c), color = "white", size = 0.2) +
  scale_fill_viridis_c(option = "inferno", direction = -1) +
  theme_minimal() +
  labs(
    title = "Mean Maximum Temperature in California",
    subtitle = "Summer 2023 (June - August)",
    fill = "Temperature (°C)"
  )
```

### Time Series for Selected Counties

```{r plot-timeseries, eval=TRUE}
# Select a few representative counties
selected_counties <- c("Los Angeles", "San Francisco", 
                      "Sacramento", "San Diego", "Fresno")

weather_subset <- weather_data %>%
  filter(county_name %in% selected_counties)

# Temperature time series
ggplot(weather_subset, aes(x = date, y = tmax_c, color = county_name)) +
  geom_line(linewidth = 1) +
  theme_minimal() +
  labs(
    title = "Daily Maximum Temperature - Selected California Counties",
    subtitle = "Summer 2023",
    x = "Date",
    y = "Max Temperature (°C)",
    color = "County"
  ) +
  theme(legend.position = "bottom")

# Precipitation time series
ggplot(weather_subset, aes(x = date, y = precip_mm, fill = county_name)) +
  geom_col(position = "dodge") +
  theme_minimal() +
  labs(
    title = "Daily Precipitation - Selected California Counties",
    subtitle = "Summer 2023",
    x = "Date",
    y = "Precipitation (mm)",
    fill = "County"
  ) +
  theme(legend.position = "bottom")
```

### Analyze Extreme Heat Days

```{r plot-heat-days, eval=TRUE}
# Map days over 35°C (95°F)
ggplot(ca_summer_map) +
  geom_sf(aes(fill = days_over_35c), color = "white", size = 0.2) +
  scale_fill_gradient2(
    low = "white",
    mid = "orange",
    high = "darkred",
    midpoint = 30
  ) +
  theme_minimal() +
  labs(
    title = "Extreme Heat Days in California",
    subtitle = "Days with maximum temperature > 35°C (Summer 2023)",
    fill = "Number of Days"
  )
```

## Step 6: H3 Hexagonal Spatial Analysis

Instead of using county boundaries, we can analyze spatial patterns using H3 hexagons - a hierarchical spatial indexing system that provides uniform hexagonal grids.

### Setup H3

```{r setup-h3}
# Install and load h3 package
library(h3jsr)
```

### Create H3 Grid for California

```{r create-h3-grid, eval=TRUE}
# Generate H3 hexagons at resolution 6 covering California
# Resolution 6 creates hexagons approximately 36 km² in area

# Ensure California counties are in WGS84
ca_counties_wgs84 <- st_transform(ca_counties_sf, 4326)

# Get the bounding box and create a simple polygon
ca_bbox <- st_bbox(ca_counties_wgs84)
bbox_poly <- st_as_sfc(ca_bbox) %>%
  st_sf() %>%
  st_set_crs(4326)

# Generate H3 cells for the bounding box
ca_h3_ids <- polygon_to_cells(bbox_poly, res = 6, simple = TRUE)

cat(paste("Generated", length(ca_h3_ids), "H3 hexagons at resolution 6\n"))

# Convert H3 cells to sf polygons and explicitly set CRS
h3_grid <- cell_to_polygon(ca_h3_ids, simple = FALSE) %>%
  st_set_crs(4326)

# Intersect with California to keep only hexagons that overlap the state
h3_grid_ca <- st_intersection(h3_grid, st_union(ca_counties_wgs84))

cat(paste("After filtering:", nrow(h3_grid_ca), "hexagons within California\n"))

# Update h3_cells to match filtered grid
h3_cells <- h3_grid_ca$h3_address

# Create centroids for data extraction
h3_centroids <- st_centroid(h3_grid_ca) %>%
  mutate(
    h3_id = h3_cells,
    lon = st_coordinates(.)[,1],
    lat = st_coordinates(.)[,2]
  ) %>%
  st_drop_geometry() %>%
  select(h3_id, lon, lat)

# Use filtered grid for visualization
h3_grid <- h3_grid_ca

head(h3_centroids)
```

```{r plot-h3-grid, eval=TRUE}
# Visualize H3 grid over California
ggplot() +
  geom_sf(data = ca_counties_sf, fill = NA, color = "gray60", size = 0.3) +
  geom_sf(data = h3_grid, fill = NA, color = "blue", size = 0.2, alpha = 0.5) +
  theme_minimal() +
  labs(
    title = "H3 Hexagonal Grid over California",
    subtitle = paste("Resolution 6 -", length(h3_cells), "hexagons")
  )
```

### Extract GridMET Data for H3 Hexagons

```{r calculate-h3-values, eval=TRUE}
# Extract temperature using H3 hexagon polygons
# This performs zonal statistics across each hexagon
tmmx_h3 <- calculate_gridmet(
  from = tmmx_processed,
  locs = h3_grid,
  locs_id = "h3_address",
  radius = 0,
  fun = "mean",
  geom = FALSE
)

# Extract precipitation using H3 hexagon polygons
pr_h3 <- calculate_gridmet(
  from = pr_processed,
  locs = h3_grid,
  locs_id = "h3_address",
  radius = 0,
  fun = "mean",
  geom = FALSE
)

head(tmmx_h3)
```

### Process and Aggregate H3 Data

```{r aggregate-h3, eval=TRUE}
# Process H3 temperature data
tmmx_h3_long <- tmmx_h3 %>%
  mutate(date = as.Date(time)) %>%
  rename(tmax_k = tmmx_0) %>%
  mutate(tmax_c = tmax_k - 273.15) %>%
  select(-time, -tmax_k)

# Process H3 precipitation data
pr_h3_long <- pr_h3 %>%
  mutate(date = as.Date(time)) %>%
  rename(precip_mm = pr_0) %>%
  select(-time)

# Merge H3 data
h3_weather_data <- tmmx_h3_long %>%
  left_join(pr_h3_long, by = c("h3_address", "date"))

# Calculate summer statistics by H3 hexagon
h3_summer_stats <- h3_weather_data %>%
  group_by(h3_address) %>%
  summarize(
    mean_tmax_c = mean(tmax_c, na.rm = TRUE),
    max_tmax_c = max(tmax_c, na.rm = TRUE),
    total_precip_mm = sum(precip_mm, na.rm = TRUE),
    days_over_35c = sum(tmax_c > 35, na.rm = TRUE),
    .groups = "drop"
  )

head(h3_summer_stats)
```

### Visualize Spatial Patterns with H3 Hexagons

```{r plot-h3-temperature, eval=TRUE}
# Join statistics with H3 grid geometry
h3_summer_map <- h3_grid %>%
  left_join(h3_summer_stats, by = "h3_address")

# Map mean temperature with H3 hexagons
ggplot(h3_summer_map) +
  geom_sf(aes(fill = mean_tmax_c), color = NA) +
  geom_sf(data = ca_counties_sf, fill = NA, color = "gray30", size = 0.3) +
  scale_fill_viridis_c(option = "inferno", direction = -1, na.value = "gray90") +
  theme_minimal() +
  labs(
    title = "Mean Maximum Temperature - H3 Hexagonal Grid",
    subtitle = "California Summer 2023 (Resolution 6)",
    fill = "Temperature (°C)"
  )
```

```{r plot-h3-precipitation, eval=TRUE}
# Map total precipitation with H3 hexagons
ggplot(h3_summer_map) +
  geom_sf(aes(fill = total_precip_mm), color = NA) +
  geom_sf(data = ca_counties_sf, fill = NA, color = "gray30", size = 0.3) +
  scale_fill_viridis_c(option = "mako", direction = -1, na.value = "gray90") +
  theme_minimal() +
  labs(
    title = "Total Precipitation - H3 Hexagonal Grid",
    subtitle = "California Summer 2023 (Resolution 6)",
    fill = "Precipitation (mm)"
  )
```

```{r plot-h3-heat-days, eval=TRUE}
# Map extreme heat days with H3 hexagons
ggplot(h3_summer_map) +
  geom_sf(aes(fill = days_over_35c), color = NA) +
  geom_sf(data = ca_counties_sf, fill = NA, color = "gray30", size = 0.3) +
  scale_fill_gradient2(
    low = "white",
    mid = "orange",
    high = "darkred",
    midpoint = 30,
    na.value = "gray90"
  ) +
  theme_minimal() +
  labs(
    title = "Extreme Heat Days - H3 Hexagonal Grid",
    subtitle = "Days > 35°C, Summer 2023 (Resolution 6)",
    fill = "Number of Days"
  )
```

::: {.callout-tip}
## Why Use H3 Hexagons?

**Advantages over administrative boundaries:**

- **Uniform sampling**: Each hexagon has equal area (~36 km² at resolution 6)
- **No edge effects**: Hexagons have equal distance to all 6 neighbors
- **Better spatial patterns**: Reveals gradients not visible with irregular county boundaries
- **Hierarchical**: Easy to aggregate/disaggregate between resolutions
- **Consistent**: Same grid system works globally

H3 hexagons provide a more uniform and statistically robust way to visualize spatial patterns in climate data.
:::

## Summary and Next Steps

In this tutorial, you've learned how to:

1. ✓ Obtain California county boundaries using `tigris`
2. ✓ Download gridMET meteorological data using `amadeus::download_data()`
3. ✓ Process raw netCDF files into usable spatial formats
4. ✓ Extract values at specific locations (county centroids)
5. ✓ Analyze and visualize temperature and precipitation patterns
6. ✓ Export results for further analysis

### Next Steps

- **Expand temporal coverage**: Download data for multiple years to analyze trends
- **Add more variables**: Explore other gridMET variables like humidity, solar radiation, or wind speed
- **Refine spatial analysis**: Use buffers or zonal statistics for more sophisticated extraction methods
- **Statistical modeling**: Use the extracted data as covariates in epidemiological or environmental models
- **Automate workflows**: Create scripts to regularly update data and generate reports

### Additional Resources

- [gridMET Documentation](https://www.climatologylab.org/gridmet.html)
- [Amadeus Package Website](https://niehs.github.io/amadeus/)
- [gridMET Variable Reference](https://www.climatologylab.org/wget-gridmet.html)
- [Terra Package Documentation](https://rspatial.github.io/terra/)

---

::: {.callout-note}
## Session Information

```{r session-info}
sessionInfo()
```
:::
